A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database

### 2022-06-08 paper/paper-jsdse/paper-jsdse.tex
incorporated nicks suggestions

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques. An example of this is the wages data made public by Singer and Willett (2003) in their book, ``Applied longitudinal data analysis,'' which can be used to teach generalized linear models, and additionally hierarchical, mixed effects, and multilevel models. The data records hourly wages of a sample of high school dropouts from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S. Department of Labor 2021a).

@@

Our approach to cleaning builds heavily on the \texttt{tidyverse} approach (Wickham et al. 2019). The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables (Wickham 2011). Tidy data shouldn't be confused with ``tame data,'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics. The resulting (tame) data is provided in a new R package called \texttt{yowie}, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

In the context of official statistics, M. van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (initial data as delivered originally), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables). What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain. In this section, we outline the steps to download the raw data (Section \ref{getdata}) and then tidy the raw data into input data, specifically for the demographic variables (Section \ref{tidydemog}) and the employment variables (Section \ref{tidyemp}), so that the resulting input data can be used downstream for validating the data as described in Section \ref{ida} and Section \ref{censor}.

@@

Finally, we save the resultant wage data on this cohort as \texttt{wages}. Note that we save \texttt{grade} variable in this dataset instead of \texttt{demog\_nlsy79} dataset because it is a longitudinal variable, while \texttt{demog\_nlsy79} is a cross-sectional dataset reflecting the state of individuals corresponding to the most recent round of the survey.

@@

In this paper, we conduct an IDA or a preliminary data analysis to assess the validity of the variable values in the cohort of data that the NLSY provides. The first step is validating numerical summaries of the raw data are the same as reported by NLSY79. This is followed by graphical summaries using methods available in \texttt{ggplot2} (Wickham 2016) and \texttt{brolgar} (Tierney, Cook, and Prvan 2020).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \texttt{id}: A unique individual's ID number.
\item
  \texttt{age\_1979}: The age of the individual in 1979.
\item
  \texttt{sex}: Sex of the individual, f = Female and m = Male.
\item
  \texttt{race} : Race of the individual belong to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
\item
  \texttt{hgc}: The highest grade completed ever.
\item
  \texttt{hgc\_i}: Integer value of the highest grade completed ever.
\item
  \texttt{hgc\_1979}: The highest grade completed in 1979 (integer value).
\item
  \texttt{ged}: Whether the individual had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  173 individuals were more than 17 years old in 1979. Thus, it looks like the description of the original data is not quite accurate that there are people older than 17 in the subset. Our decision is to also include them in the refreshed data as the new data contains an age variable, so analysts could filter them later.
\item
  715 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because:

@@

The reporting of highest grade completed is confusing. There are several ways this is reported, including the highest grade ever completed and also the \texttt{hgc} at each survey year. To match the original data, it is appropriate to use the \texttt{hgc} while in high school. The documentation suggests this variable is available, but it is actually not. Hence, to match the original data, we have calculated the \texttt{hgc} to match the \texttt{hgc} achieved in the years between 1979 and 1994 based on the yearly survey value. The result does not exactly match the original for a few individuals.

@@

\begin{itemize}
\tightlist
\item
  It primarily illustrates the steps in cleaning and processing a well-known longitudinal data so that it can be further refreshed as a textbook data set as new data emerges. Choices made during the cleaning and processing can affect findings made with the data, are transparent, and can be a basis for classroom discussions.
\item
  Some of the methods shown, and lessons learned about the data, can also be considered generally for data cleaning, initial data analysis and exploratory data analysis, useful for working with other data sets.
\item
  The refreshed data can be used for teaching advanced linear modeling, using almost identical code from Singer and Willett (2003), and thus make those lessons more current for today's students.
\item
  It provides the data to develop a case study for teaching exploratory data analysis of longitudinal data, focusing on the individual experience. The vignettes of the \texttt{brolgar} (Tierney, Cook, and Prvan 2020) package provide useful guides for the case study content.
\end{itemize}

>>>>

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples on how to apply different techniques. An example of this is the wages data made public by Singer and Willett (2003) in their book, ``Applied longitudinal data analysis,'' which can be used to teach generalized linear models, and additionally hierarchical, mixed effects, and multilevel models. The data records hourly wages of a sample of high school dropouts from 1979-1994, along with demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S. Department of Labor 2021a).

@@

Our approach to cleaning builds heavily on the \texttt{tidyverse} approach (Wickham et al. 2019). The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables (Wickham 2011). Tidy data shouldn't be confused with ``tame data,'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics. The resulting (tame) data is provided in a new R package called \texttt{yowie}, which includes the code so the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

In the context of official statistics, M. van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (initial data as delivered originally), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables). What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain. In this section, we outline the steps to download the raw data (Section \ref{getdata}) and then tidy the raw data into input data, specifically for the demographic variables (Section \ref{tidydemog}) and the employment variables (Section \ref{tidyemp}), so the resulting input data can be used downstream for validating the data as described in Section \ref{ida} and Section \ref{censor}.

@@

Finally, we save the resultant wage data on this cohort as \texttt{wages}. Note that we save the \texttt{grade} variable in this dataset instead of the \texttt{demog\_nlsy79} dataset because it is a longitudinal variable, while \texttt{demog\_nlsy79} is a cross-sectional dataset reflecting the state of individuals corresponding to the most recent round of the survey.

@@

In this paper, we conduct an IDA or a preliminary data analysis to assess the validity of the variable values in the cohort of data that the NLSY provides. The first step is validating that the numerical summaries of the raw data are the same as reported by NLSY79. This is followed by graphical summaries using methods available in \texttt{ggplot2} (Wickham 2016) and \texttt{brolgar} (Tierney, Cook, and Prvan 2020).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \texttt{id}: A unique individual's ID number.
\item
  \texttt{age\_1979}: The age of the individual in 1979.
\item
  \texttt{sex}: Sex of the individual, f = Female and m = Male.
\item
  \texttt{race} : Race the individual belongs to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
\item
  \texttt{hgc}: The highest grade completed ever.
\item
  \texttt{hgc\_i}: Integer value of the highest grade completed ever.
\item
  \texttt{hgc\_1979}: The highest grade completed in 1979 (integer value).
\item
  \texttt{ged}: Whether the individual had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  173 individuals were more than 17 years old in 1979. Thus, it looks like the description of the original data, that there are people older than 17 in the subset, is not quite accurate. Our decision is to also include them in the refreshed data as the new data contains an age variable, so analysts could filter them later.
\item
  715 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because:

@@

The variables from ``highest grade completed'' have some confusion. There are several ways this is reported, including the highest grade ever completed and also the \texttt{hgc} at each survey year. To match the original data, it is appropriate to use the \texttt{hgc} while in high school. The documentation suggests this variable is available, but it is actually not. Hence, to match the original data, we have calculated the \texttt{hgc} to match the \texttt{hgc} achieved in the years between 1979 and 1994 based on the yearly survey value. The result does not exactly match the original for a few individuals.

@@

\begin{itemize}
\tightlist
\item
  It primarily illustrates the steps in cleaning and processing a well-known longitudinal data so that it can be further refreshed as a textbook data set as new data emerges. Choices made during the cleaning and processing can affect findings made with the data, are transparent, and can be a basis for classroom discussions.
\item
  Some of the methods shown, and lessons learned about the data, can also be considered to be useful for working with other data sets, and also apply generally for data cleaning, initial data analysis and exploratory data analysis.
\item
  The refreshed data can be used for teaching advanced linear modeling, using almost identical code from Singer and Willett (2003), and thus make those lessons more current for today's students.
\item
  It provides the data to develop a case study for teaching exploratory data analysis of longitudinal data, focusing on the individual experience. The vignettes of the \texttt{brolgar} (Tierney, Cook, and Prvan 2020) package provide useful guides for the case study content.
\end{itemize}

### 2022-06-08 paper/paper.Rmd
incorporated nicks suggestions

Our approach to cleaning builds heavily on the `tidyverse` approach [@tidyverse]. The data is first organized into "tidy data" [@WickhamHadley2014TD] and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables [@plyr]. Tidy data shouldn't be confused with "tame data," which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics. The resulting (tame) data is provided in a new R package called `r yowie`, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

Finally, we save the resultant wage data on this cohort as `wages`. Note that we save `grade` variable in this dataset instead of `demog_nlsy79` dataset because it is a longitudinal variable, while `demog_nlsy79` is a cross-sectional dataset reflecting the state of individuals corresponding to the most recent round of the survey.

@@

  1. `id`: A unique individual's ID number.
  2. `age_1979`: The age of the individual in 1979.
  3. `sex`: Sex of the individual, f = Female and m = Male.
  4. `race` : Race of the individual belong to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
  5. `hgc`: The highest grade completed ever.
  6. `hgc_i`: Integer value of the highest grade completed ever.
  7. `hgc_1979`: The highest grade completed in 1979 (integer value). 
  8. `ged`: Whether the individual had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.

@@

The reporting of highest grade completed is confusing. There are several ways this is reported, including the highest grade ever completed and also the `hgc` at each survey year. To match the original data, it is appropriate to use the `hgc` while in high school. The documentation suggests this variable is available, but it is actually not. Hence, to match the original data, we have calculated the `hgc` to match the `hgc` achieved in the years between 1979 and 1994 based on the yearly survey value. The result does not exactly match the original for a few individuals.

@@

- It primarily illustrates the steps in cleaning and processing a well-known longitudinal data so that it can be further refreshed as a textbook data set as new data emerges. Choices made during the cleaning and processing can affect findings made with the data, are transparent, and can be a basis for classroom discussions.
- Some of the methods shown, and lessons learned about the data, can also be considered generally for data cleaning, initial data analysis and exploratory data analysis, useful for working with other data sets.
- The refreshed data can be used for teaching advanced linear modeling, using almost identical code from @SingerJudithD2003Alda, and thus make those lessons more current for today's students.
- It provides the data to develop a case study for teaching exploratory data analysis of longitudinal data, focusing on the individual experience. The vignettes of the `brolgar` [@brolgar] package provide useful guides for the case study content.

>>>>

Our approach to cleaning builds heavily on the `tidyverse` approach [@tidyverse]. The data is first organized into "tidy data" [@WickhamHadley2014TD] and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables [@plyr]. Tidy data shouldn't be confused with "tame data," which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics. The resulting (tame) data is provided in a new R package called `r yowie`, which includes the code so the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

Finally, we save the resultant wage data on this cohort as `wages`. Note that we save the `grade` variable in this dataset instead of the `demog_nlsy79` dataset because it is a longitudinal variable, while `demog_nlsy79` is a cross-sectional dataset reflecting the state of individuals corresponding to the most recent round of the survey.

@@

  1. `id`: A unique individual's ID number.
  2. `age_1979`: The age of the individual in 1979.
  3. `sex`: Sex of the individual, f = Female and m = Male.
  4. `race` : Race the individual belongs to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
  5. `hgc`: The highest grade completed ever.
  6. `hgc_i`: Integer value of the highest grade completed ever.
  7. `hgc_1979`: The highest grade completed in 1979 (integer value). 
  8. `ged`: Whether the individual had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.

@@

The variables from "highest grade completed" have some confusion. There are several ways this is reported, including the highest grade ever completed and also the `hgc` at each survey year. To match the original data, it is appropriate to use the `hgc` while in high school. The documentation suggests this variable is available, but it is actually not. Hence, to match the original data, we have calculated the `hgc` to match the `hgc` achieved in the years between 1979 and 1994 based on the yearly survey value. The result does not exactly match the original for a few individuals.

@@

- It primarily illustrates the steps in cleaning and processing a well-known longitudinal data so that it can be further refreshed as a textbook data set as new data emerges. Choices made during the cleaning and processing can affect findings made with the data, are transparent, and can be a basis for classroom discussions.
- Some of the methods shown, and lessons learned about the data, can also be considered to be useful for working with other data sets, and also apply generally for data cleaning, initial data analysis and exploratory data analysis.
- The refreshed data can be used for teaching advanced linear modeling, using almost identical code from @SingerJudithD2003Alda, and thus make those lessons more current for today's students.
- It provides the data to develop a case study for teaching exploratory data analysis of longitudinal data, focusing on the individual experience. The vignettes of the `brolgar` [@brolgar] package provide useful guides for the case study content.
